---
title: "COVID-19 Data Report"
author: "Ayush Jain"
date: "2025-06-18"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# COVID 19 Data Report

This report analyzes COVID-19 data from 2020 through 2023. The dataset, compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, documents the impact of COVID-19 across the United States and globally.

Data Source: - https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series

## Library Imports

```{r library, results='hide', warning=FALSE, message=FALSE}
library(readr)
library(tidyverse)
library(party)
library(caret)
library(e1071) 
library(usmap)
library(ggplot2)
library(rworldmap)
library(plotly)
library(nnet)
library(usmap)
library(ggnewscale)
```

## Importing the data

The data are sourced directly from publicly available GitHub CSV files maintained by Johns Hopkins University.

```{r data load}
url_in = 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/'
filesnames = c('time_series_covid19_confirmed_US.csv',
               'time_series_covid19_confirmed_global.csv',
               'time_series_covid19_deaths_US.csv',
               'time_series_covid19_deaths_global.csv',
               'time_series_covid19_recovered_global.csv'
               )
urls = str_c(url_in, filesnames)

us_cases = read_csv(urls[1])
global_cases = read_csv(urls[2])
us_deaths = read_csv(urls[3])
global_deaths = read_csv(urls[4])
global_recovered = read_csv(urls[5])

uid_lookup_url = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid = read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

## Data Cleaning and Transformation

Unnecessary columns are removed and data is reshaped to facilitate time series and spatial analyses. Feature engineering is used to derive new summary statistics.

```{r data transform}
global_cases = global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', 'Lat', 'Long'),
               names_to='date',
               values_to='cases') %>%
  select(-c(Lat, Long))

global_deaths = global_deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', 'Lat', 'Long'),
               names_to='date',
               values_to='deaths') %>%
  select(-c(Lat, Long))

global_recovered = global_recovered %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', 'Lat', 'Long'),
               names_to='date',
               values_to='recovered') %>%
  select(-c(Lat, Long))

us_cases = us_cases %>%
  pivot_longer(cols=-(UID:Combined_Key),
               names_to='date',
               values_to='cases') %>%
  select(Admin2:cases) %>%
  mutate(date=mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths = us_deaths %>%
  pivot_longer(cols=-(UID:Population),
               names_to='date',
               values_to='deaths') %>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date)) %>%
  select(-c(Lat, Long_))
```

Data tables are joined to combine cases, deaths, and recovered counts.

```{r data join}
global = global_cases %>%
  full_join(global_deaths) %>%
  full_join(global_recovered) %>%
  rename(Country_Region='Country/Region',
         Province_State ='Province/State') %>%
  mutate(date=mdy(date))

global = global %>%
  filter(cases>0)

us = us_cases %>%
  full_join((us_deaths))
```
Further join and cleanup for global data

```{r data join for uid}

global = global %>%
  unite('Combined_Key',
       c(Province_State, Country_Region),
       sep=",",
       na.rm=TRUE,
       remove=FALSE)

global = global %>%
  left_join(uid, by=c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, recovered, Population, Combined_Key)
```

Aggregate at the US state and national levels, and compute new daily cases/deaths

```{r data cleaning for us state data, warning=FALSE}
us_by_state = us %>% 
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths),
            population=sum(Population)) %>%
  mutate(deaths_per_mill=deaths*1000000/population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, population) %>%
  ungroup()

us_totals = us_by_state %>% 
  group_by(Country_Region, date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths),
            population=sum(population)) %>%
  mutate(deaths_per_mill=deaths*1000000/population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, population) %>%
  ungroup()

max(us_by_state$date)
min(us_by_state$date)

us_by_state =  us_by_state %>% 
  mutate(new_cases = cases-lag(cases),
         new_deaths=deaths-lag(deaths))

us_totals =  us_totals %>% 
  mutate(new_cases = cases-lag(cases),
         new_deaths= deaths-lag(deaths))

tail(us_totals %>% select(new_cases, new_deaths, everything()))
```

Summarize global COVID-19 statistics at the country level

```{r data cleaning for global data, warning=FALSE}
global_totals = global %>% 
  group_by(Country_Region, date) %>%
  summarize(cases=sum(cases), deaths=sum(deaths),
            recovered=sum(recovered),
            population=sum(Population)) %>%
  mutate(deaths_per_mill=deaths*1000000/population) %>%
  select(Country_Region, date, cases, deaths, recovered, deaths_per_mill, population) %>%
  ungroup()

global_totals =  global_totals %>% 
  mutate(new_cases = cases-lag(cases),
         new_deaths=deaths-lag(deaths),
         new_recovered=recovered-lag(recovered))

global_totals = global_totals %>%
  group_by(Country_Region) %>%
  summarize(deaths=max(deaths), 
            cases=max(cases),
            recovered=max(recovered),
            population=max(population),
            active=cases-recovered,
            cases_per_thou = 1000*cases/population,
            deaths_per_thou = 1000*deaths/population,
            recovered_per_thou = 1000*recovered/population) %>%
  filter(cases>0, population>0)
```

Identify countries with highest/lowest death and recovery rates. It also reflects the country with maximum active cases.

```{r global data summary}
global_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

global_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

global_totals %>%
  slice_min(recovered_per_thou, n=10) %>%
  select(recovered_per_thou, cases_per_thou, everything())

global_totals %>%
  slice_max(recovered_per_thou, n=10) %>%
  select(recovered_per_thou, cases_per_thou, everything())

global_totals %>%
  slice_max(active, n=10) %>%
  select(active, cases_per_thou, everything())
```

## Data Visualization

### COVID-19 Cases and Deaths in the US Over Time

The following plot shows cumulative COVID-19 cases and deaths over time in the United States.

```{r us total case and death plot}
us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = cases, color = "Cases")) +
  geom_point(aes(y = cases, color = "Cases")) +
  geom_line(aes(y = deaths, color = "Deaths")) +
  geom_point(aes(y = deaths, color = "Deaths")) +
  scale_y_log10() +
  labs(title = "Cumulative COVID-19 Cases and Deaths in the US",
       y = "Count (log scale)",
       color = "Metric") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1))
```

### New Cases and Deaths in the US Over Time

This plot shows new COVID-19 cases and deaths per day in the US.

```{r us new cases and death plot, warning=FALSE}
us_totals %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = new_cases, color = "New Cases")) +
  geom_point(aes(y = new_cases, color = "New Cases")) +
  geom_line(aes(y = new_deaths, color = "New Deaths")) +
  geom_point(aes(y = new_deaths, color = "New Deaths")) +
  scale_y_log10() +
  labs(title = "Daily New COVID-19 Cases and Deaths in the US",
       y = "Count (log scale)",
       color = "Metric") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1))

```

US States: Cases and Deaths Per Thousand - Aggregate cases and deaths per thousand by US state. Display states with highest and lowest death rates.

```{r us state total data}
us_state_totals = us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths=max(deaths), cases=max(cases),
            population=max(population),
            cases_per_thou = 1000*cases/population,
            deaths_per_thou = 1000*deaths/population) %>%
  filter(cases>0, population>0)
  
us_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

us_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

```

### Total COVID-19 Cases in US States

This map shows total reported cases by US state. California has reported the maximum cases in the country.

```{r us state total cases spatial data}
us_state_plot_data = us_state_totals %>%
  select(Province_State, cases) %>%
  rename(state = "Province_State")

plot_usmap(data = us_state_plot_data,
           values = "cases",
           color = "blue",
           labels = TRUE) +
  scale_fill_continuous(name = "Total Cases", label = scales::comma) +
  labs(title = "Total COVID-19 Cases by US State") +
  theme(panel.background = element_rect(color = "black", fill = "lightblue"))
```

### Total COVID-19 Deaths in US States

This map shows total deaths reported in each US state. The below spatial analysis shows that most number of deaths have been reported in California followed by Texas. The least number of deaths are reported in American Samoa.

```{r us state deaths spatial data}
us_state_deaths_plot_data = us_state_totals %>%
  select(Province_State, deaths) %>%
  rename(state = "Province_State")

plot_usmap(data = us_state_deaths_plot_data,
           values = "deaths",
           color = "red",
           labels = TRUE) +
  scale_fill_continuous(name = "Total Deaths", label = scales::comma) +
  labs(title = "Total COVID-19 Deaths by US State") +
  theme(panel.background = element_rect(color = "black", fill = "lightblue"))
```

### Global COVID-19 Cases by Country

The following map visualizes total COVID-19 cases per country.

```{r global total cases spatial data, warning=FALSE}
global_plot_data = global_totals %>%
  select(Country_Region, cases) %>%
  rename(country = "Country_Region")

CASES <- global_plot_data %>%
  group_by(country) %>%
  summarise(`Total cases` = sum(cases)) %>%
  mutate(country = gsub("_", " ", country))

COVID.map <- joinCountryData2Map(CASES, joinCode = "NAME", nameJoinColumn = "country")
par(mar = c(0, 0, 1, 0))
mapCountryData(COVID.map, nameColumnToPlot = "Total cases", colourPalette = 'topo')
```

### Global COVID-19 Deaths by Country

The below graph shows the deaths caused by Covid 19 for all of the countries in the world.

```{r global deaths spatial data, warning=FALSE}
global_deaths_plot_data = global_totals %>%
  select(Country_Region, deaths) %>%
  rename(country = "Country_Region")

CASES <- global_deaths_plot_data %>%
  group_by(country) %>%
  summarise(`Total deaths` = sum(deaths)) %>%
  mutate(country = gsub("_", " ", country))

COVID.map <- joinCountryData2Map(CASES, joinCode = "NAME", nameJoinColumn = "country")
par(mar = c(0,0,1,0))
mapCountryData(COVID.map, nameColumnToPlot = "Total deaths", colourPalette = 'heat')
```

### Global Recoveries

The below graph shows the recoveries from Covid 19 for all of the countries in the world.

```{r global recoveries spatial data, warning=FALSE}
global_recovered_plot_data = global_totals %>%
  select(Country_Region, recovered) %>%
  rename(country = "Country_Region")

CASES <- global_recovered_plot_data %>%
  group_by(country) %>%
  summarise(`Total recovered` = sum(recovered)) %>%
  mutate(country = gsub("_", " ", country))

COVID.map <- joinCountryData2Map(CASES, joinCode = "NAME", nameJoinColumn = "country")
par(mar = c(0,0,1,0))
mapCountryData(COVID.map, nameColumnToPlot = "Total recovered", colourPalette = 'terrain')
```

### Global Active Cases

The below graph shows the total active cases of Covid 19 for all of the countries in the world.

```{r global active cases spatial data, warning=FALSE}
global_active_plot_data = global_totals %>%
  select(Country_Region, active) %>%
  rename(country = "Country_Region")

CASES <- global_active_plot_data %>%
  group_by(country) %>%
  summarise(`Total active` = sum(active)) %>%
  mutate(country = gsub("_", " ", country))

COVID.map <- joinCountryData2Map(CASES, joinCode = "NAME", nameJoinColumn = "country")
par(mar = c(0,0,1,0))
mapCountryData(COVID.map, nameColumnToPlot = "Total active", colourPalette = 'rainbow')
```

## Modelling

The COVID 19 data is modeled to predict the number of deaths with relation to all of the cases reported in us and world. We use linear regression, SVM, and neural network models to predict COVID-19 death rates based on cases and recoveries, comparing model performance.

### Linear Regression

Predicting deaths per thousand using cases per thousand (US state-level data). Linear Regression tries to use one or more factors to identify linear relationship with the resultant value.

```{r single linear regression for us}
mod = lm(deaths_per_thou ~ cases_per_thou, data=us_state_totals)
summary(mod)
```

The below graph shows the linear regression for us state data. It compares the actual data points and the ones predicted using the model.

```{r single linear regression for us plot}
us_total_w_pred = us_state_totals %>% 
  mutate(pred=predict(mod))

us_total_w_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  labs(
    title = "Linear Regression: Observed vs. Predicted Deaths per Thousand (US States)",
    x = "Cases per Thousand",
    y = "Deaths per Thousand"
  ) +
  theme_minimal()
```

The below Linear regression model predicts the death rate using the total cases for all the countries across the globe. The graph shows the predicted and actual values for global dataset.

```{r single linear regression for world}
mod = lm(deaths_per_thou ~ cases_per_thou, data=global_totals)
summary(mod)

global_totals_w_pred = global_totals %>% 
  mutate(pred=predict(mod))

global_totals_w_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  labs(
    title = "Linear Regression: Observed vs. Predicted Deaths per Thousand (Global)",
    x = "Cases per Thousand",
    y = "Deaths per Thousand"
  ) +
  theme_minimal()
```

#### Multiple Linear Regression with Cases and Recovery and 3D Plot for Predicted and actual values

The below Linear regression model predicts the death rate using the total cases and recovery rate for all the countries accross the globe. It also shows the 3 dimensional graphs where the predicted and actual values are shown in 2 different graphs.


```{r multiple linear regression for world}
mod = lm(deaths_per_thou ~ cases_per_thou +  recovered_per_thou, data=global_totals)
summary(mod)

global_totals_w_pred = global_totals %>% 
  mutate(pred=predict(mod))

global_totals_w_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red") +
  labs(
    title = "Multiple Linear Regression: Deaths per Thousand vs. Cases and Recovery",
    x = "Cases per Thousand",
    y = "Deaths per Thousand"
  ) +
  theme_minimal()

plot_ly(data = global_totals_w_pred, x = ~cases_per_thou, y = ~recovered_per_thou, z = ~pred,
        type = 'scatter3d', mode = 'markers',
        marker = list(size = 5, color = ~pred, colorscale = 'Viridis')) %>%
  layout(scene = list(xaxis = list(title = 'Cases per Thousand'),
                      yaxis = list(title = 'Recovered Per Thousand'),
                      zaxis = list(title = 'Predicted Deaths per Thousand')))

plot_ly(data = global_totals_w_pred, x = ~cases_per_thou, y = ~recovered_per_thou, z = ~deaths_per_thou,
        type = 'scatter3d', mode = 'markers',
        marker = list(size = 5, color = ~deaths_per_thou, colorscale = 'Viridis')) %>%
  layout(scene = list(xaxis = list(title = 'Cases per Thousand'),
                      yaxis = list(title = 'Recovered Per Thousand'),
                      zaxis = list(title = 'Deaths per Thousand')))
```

Model Performance Metrics: The below graph shows the performance of the model. It reflects the actual and predicted values with a Root Mean Squared Value of 1.14

``` {r multiple linear regression for world performance}
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}
actual_values <- global_totals_w_pred$deaths_per_thou
predicted_values <- global_totals_w_pred$pred

print(paste("RMSE:", rmse(actual_values, predicted_values)))
print(paste("R2:", caret::R2(predicted_values, actual_values)))

plot(actual_values, type = "l", col = "blue", main = "Actual vs Predicted Values", xlab = "Index", ylab = "Deaths per Thousand")
lines(predicted_values, col = "red")
legend("topright", legend = c("Actual", "Predicted"), col = c("blue", "red"), lty = 1)
```

### SVM

The below code uses SVM model using the total cases and recoveries for covid 19 for global data. The performance of this model is not good as compared to Linear Regression because of high RMSE.

```{r svm for world}
regressor <- svm(deaths_per_thou ~ cases_per_thou + recovered_per_thou,
                 data = global_totals,
                 kernel = "linear", cost = 10, scale = FALSE)
predictions <- predict(regressor, newdata = global_totals)

print(paste("SVM RMSE:", rmse(global_totals$deaths_per_thou, predictions)))

plot(global_totals$deaths_per_thou, type = "l", col = "blue", main = "SVM: Actual vs Predicted", xlab = "Index", ylab = "Deaths per Thousand")
lines(predictions, col = "red")
legend("topright", legend = c("Actual", "Predicted"), col = c("blue", "red"), lty = 1)
```

### Neural Network

The below code uses neural network model using the total cases and recoveries for covid 19 for global data. The performance of this model is best as it has the lowest Root Mean Squared value among the models tested as part of this report.

```{r nnet for world}
nn = nnet(deaths_per_thou ~ cases_per_thou + recovered_per_thou, data = global_totals, size = 10, maxit = 1000, linout = TRUE)
predictions <- predict(nn, newdata = global_totals)

print(paste("Neural Network RMSE:", rmse(global_totals$deaths_per_thou, predictions)))

plot(global_totals$deaths_per_thou, type = "l", col = "blue", main = "Neural Network: Actual vs Predicted", xlab = "Index", ylab = "Deaths per Thousand")
lines(predictions, col = "red")
legend("topright", legend = c("Actual", "Predicted"), col = c("blue", "red"), lty = 1)
```

## Bias
While the data from Johns Hopkins University aggregates multiple sources, some cases may go unreported due to limitations in data availability and cross-country reporting standards. Data coverage is more comprehensive for the US compared to other countries, which could introduce bias.

On the analysis side, variable selection (cases, deaths, recovered) and certain model choices reflect analyst bias in focusing on specific aspects of the pandemic. Visualization and mapping may omit regions due to missing identifiers.

## Conclusion

This report provides an overview and modeling of COVID-19 incidence and mortality trends both in the US and globally. Countries and US states with the highest and lowest case and death rates were identified. Neural Network models offered the best predictive accuracy for deaths per thousand among models tested. The analysis highlights patterns that can assist with public health planning and response, but should always be interpreted with awareness of data and modeling limitations.